The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2009-08-19 script/en/chapters/primes.tex
CT script: Sage parts finished in de + en; several corrections
CT script: Minor corrections, de + en; final now for Beta04
CT script: Minor corrections, mostly layout, de + en; final now for CT 1.4.30 Beta04

$e_9, \cdots, e_{17}$ are also composite, which means that this formula is not
particularly useful.

Remark: However, what is special about these numbers is that any pair of 
them does not have a common factor other than $1$\footnote{%
This can easily be shown via the following {\em greatest common divisor} ($gcd$) rule $gcd(a,b) = gcd(b-\lfloor b/a \rfloor,a)$ (see page \pageref{NumberTheory_Appendix_GCD}):
We have for $i<j$: \\
$gcd(e_i,e_j) \le gcd(e_1 \cdots e_i \cdots e_{j-1}, e_j) = gcd(e_j - e_1 \cdots e_i \cdots e_{j-1}, e_1 \cdots e_i \cdots e_{j-1}) 
= gcd(1, e_1 \cdots e_i \cdots e_{j-1}) = 1$.
}. Therefore they are \index{Prime number!relative prime}\index{Relatively prime} {\em relatively prime}.

% --------------------------------------------------------------------------
% \pagebreak %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vskip +10 pt
\subsection{\texorpdfstring{$f(n) = n^2 + n + 41$}{f(n) = n\^{}2 + n + 41}}
\label{L-Polynomfunktion01-41}
   This sequence starts off very {\em promisingly},  
   but is far from being a proof.

$$
 \begin{array}{lll}
f(0) = 41 & & \mapsto {\rm ~prime} \\
f(1) = 43 & & \mapsto {\rm ~prime} \\
f(2) = 47 & & \mapsto {\rm ~prime} \\
f(3) = 53 & & \mapsto {\rm ~prime} \\
f(4) = 61 & & \mapsto {\rm ~prime} \\
f(5) = 71 & & \mapsto {\rm ~prime} \\
f(6) = 83 & & \mapsto {\rm ~prime} \\
f(7) = 97 & & \mapsto {\rm ~prime} \\
\vdots \\
f(33) = 1,163 & & \mapsto {\rm ~prime} \\
f(34) = 1,231 & & \mapsto {\rm ~prime} \\
f(35) = 1,301 & & \mapsto {\rm ~prime} \\
f(36) = 1,373 & & \mapsto {\rm ~prime} \\
f(37) = 1,447 & & \mapsto {\rm ~prime} \\
f(38) = 1,523 & & \mapsto {\rm ~prime} \\
f(39) = 1,601 & & \mapsto {\rm ~prime} \\
f(40) = 1681 & = 41 \cdot 41 & \mapsto {\rm ~NOT~prime}! \\
f(41) = 1763 & = 41 \cdot 43 & \mapsto {\rm ~NOT~prime}! \\
\end{array} 
$$

@@

% --------------------------------------------------------------------------
\pagebreak %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%   be_2005
% \vskip +10 pt
\subsection{\texorpdfstring{$f(n) = n^2 - 79 \cdot n + 1,601$}{f(n) = n\^{}2 - 79 ... n + 1,601}}
\label{L-Polynomfunktion02-1601}   \label{SrcPrimArith1}
   This function\footnote{%
   See \hyperlink{Primes_Appendix_A}{Appendix A of this chapter} for the source code
   to compute the table using Sage\index{Sage}.}  %Pari-GP.\index{Pari-GP}}
   delivers prime numbers for all values from $n=0$ to $n=79$.
Unfortunately $f(80) = 1,681 = 11 \cdot 151$ is not a prime number. To this
date, no function has been found that delivers more prime numbers in a row. On
the other hand, each prime occurs twice (first in the decreasing then in the
increasing sequence), which means that the algorithm delivers a total of 40
different prime values (these are the same ones as delivered by the function
in chapter~\ref{L-Polynomfunktion01-41})\footnote{%
   Another quadratic polynom, which delivers these primes, is: 
   $f(n) = n^2 - 9 \cdot n + 61$.\\
   Among the first 1000 sequence elements more than 50\% are prime
   (See \hyperlink{Primes_Appendix_A}{Appendix A of this chapter}).}.
$$
\begin{array}{|ll||ll|}
\hline
f(0)  = 1.601 & \mapsto {\rm ~prim}   &   f(26) =   223 & \mapsto {\rm ~prim} \\
f(1)  = 1.523 & \mapsto {\rm ~prim}   &   f(27) =   197 & \mapsto {\rm ~prim} \\
f(2)  = 1.447 & \mapsto {\rm ~prim}   &   f(28) =   173 & \mapsto {\rm ~prim} \\
f(3)  = 1.373 & \mapsto {\rm ~prim}   &   f(29) =   151 & \mapsto {\rm ~prim} \\
f(4)  = 1.301 & \mapsto {\rm ~prim}   &   f(30) =   131 & \mapsto {\rm ~prim} \\
f(5)  = 1.231 & \mapsto {\rm ~prim}   &   f(31) =   113 & \mapsto {\rm ~prim} \\
f(6)  = 1.163 & \mapsto {\rm ~prim}   &   f(32) =    97 & \mapsto {\rm ~prim} \\
f(7)  = 1.097 & \mapsto {\rm ~prim}   &   f(33) =    83 & \mapsto {\rm ~prim} \\
f(8)  = 1.033 & \mapsto {\rm ~prim}   &   f(34) =    71 & \mapsto {\rm ~prim} \\
f(9)  =   971 & \mapsto {\rm ~prim}   &   f(35) =    61 & \mapsto {\rm ~prim} \\
f(10) =   911 & \mapsto {\rm ~prim}   &   f(36) =    53 & \mapsto {\rm ~prim} \\
f(11) =   853 & \mapsto {\rm ~prim}   &   f(37) =    47 & \mapsto {\rm ~prim} \\
f(12) =   797 & \mapsto {\rm ~prim}   &   f(38) =    43 & \mapsto {\rm ~prim} \\
f(13) =   743 & \mapsto {\rm ~prim}   &   f(39) =    41 & \mapsto {\rm ~prim} \\
f(14) =   691 & \mapsto {\rm ~prim}   &   f(40) =    41 & \mapsto {\rm ~prim} \\              
f(15) =   641 & \mapsto {\rm ~prim}   &   f(41) =    43 & \mapsto {\rm ~prim} \\              
f(16) =   593 & \mapsto {\rm ~prim}   &   f(42) =    47 & \mapsto {\rm ~prim} \\              
f(17) =   547 & \mapsto {\rm ~prim}   &   f(43) =    53 & \mapsto {\rm ~prim} \\              
f(18) =   503 & \mapsto {\rm ~prim}   &   \cdots  &  \\                                       
f(19) =   461 & \mapsto {\rm ~prim}   &   f(77) = 1.447 & \mapsto {\rm ~prim} \\             
f(20) =   421 & \mapsto {\rm ~prim}   &   f(78) = 1.523 & \mapsto {\rm ~prim} \\             
f(21) =   383 & \mapsto {\rm ~prim}   &   f(79) = 1.601 & \mapsto {\rm ~prim} \\             
f(22) =   347 & \mapsto {\rm ~prim}   &   f(80) = 41 \cdot 41 & \mapsto {\rm ~NOT~prim!} \\
f(21) =   383 & \mapsto {\rm ~prim}   &   f(81) = 41 \cdot 43 & \mapsto {\rm ~NOT~prim!} \\ 
f(22) =   347 & \mapsto {\rm ~prim}   &   f(82) = 1.847 & \mapsto {\rm ~prim} \\             
f(23) =   313 & \mapsto {\rm ~prim}   &   f(83) = 1.933 & \mapsto {\rm ~prim} \\              
f(24) =   281 & \mapsto {\rm ~prim}   &   f(84) = 43 \cdot 47 &  \mapsto {\rm ~NOT~prim!} \\
f(25) =   251 & \mapsto {\rm ~prim}   &   & \\
\hline
\end{array} 
$$

% --------------------------------------------------------------------------
\vskip +10 pt
\subsection[Polynomial functions 
    \texorpdfstring{$f(x) = a_n x^n + a_{n-1}x^{n-1} + \cdots + a_1 x^1 + a_0$}
                   {f(x) = a\_n x\^{}n + a\_n-1 x\^{}n-1 + ... + a\_1 x\^{}1 + a\_0}]
    {Polynomial functions\index{Polynomial} 
    $f(x) = a_n x^n + a_{n-1}x^{n-1} + \cdots + a_1 x^1 + a_0$ ~ 
    ($a_i$ in ${\mathbb Z}$, $n \geq 1$)}\index{Polynomial}

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\subsection[Catalan's conjecture]{Catalan's conjecture\footnotemark}
    \footnotetext{%
Eugene Charles Catalan, Belgian mathematician, May 5, 1814$-$Feb 14, 1894.\\
After him the so-called {\em Catalan numbers}
$A(n) = (1 / (n+1) ) * (2n)! / (n!)^2 $ \\
$=  1, 2, 5, 14, 42, 132, 429, 1430, 
4862, 16796, 58786, 208012, 742900, 2674440, 9694845, ... $
are named.
    }
\begin{sloppypar}
Catalan \index{Catalan Eugene}  \index{Number!Catalan}
conjectured that $ C_4 \;$ is a prime:
$$
 \begin{array}{l}
C_0 = 2, \\
C_1 = 2^{C_0} - 1,  \\
C_2 = 2^{C_1} - 1,  \\
C_3 = 2^{C_2} - 1, \\
C_4 = 2^{C_3} - 1, \cdots \\
\end{array}
$$
(see \url{http://www.utm.edu/research/primes/mersenne.shtml} under Conjectures and
Unsolved Problems).
\end{sloppypar}

@@

\vskip +12pt Web links:\\
{\href{http://www.math.Princeton.EDU/~arbooker/nthprime.html}
      {http://www.math.Princeton.EDU/\~{}arbooker/nthprime.html}.} \\
{\href{http://www.utm.edu/research/primes/notes/by_year.html}
      {\tt http://www.utm.edu/research/primes/notes/by\_year.html}.}

@@

% ---------------------------------------------------------------------------
% \newpage
\subsection{Check primality of integers generated by quadratic functions}

@@

\noindent The last two lines of the output contain a small statistics.
You can see that $f(n)$ generates
48 primes when $0 \leq n \leq 50$, which is approximately 94\% of the
values generated by $f(n)$ for $n = 0, 1, 2, 3, \dots, 50$.\\

>>>>

\noindent Also $e_9, \cdots, e_{17}$ are composite, which means that this formula is not
particularly useful.

\begin{remark}{:}\\
However, what is special about these numbers is that any pair of 
them does not have a common factor other than $1$\footnote{%
This can easily be shown via the following {\em greatest common divisor} ($gcd$) rule $gcd(a,b) = gcd(b-\lfloor b/a \rfloor,a)$ (see page \pageref{NumberTheory_Appendix_GCD}):
We have for $i<j$: \\
$gcd(e_i,e_j) \le gcd(e_1 \cdots e_i \cdots e_{j-1}, e_j) = gcd(e_j - e_1 \cdots e_i \cdots e_{j-1}, e_1 \cdots e_i \cdots e_{j-1}) 
= gcd(1, e_1 \cdots e_i \cdots e_{j-1}) = 1$.
}. Therefore they are \index{Prime number!relative prime}\index{Relatively prime}
{\em relatively prime}.
\end{remark}

% --------------------------------------------------------------------------
\vskip +10 pt
\subsection{\texorpdfstring{$f(n) = n^2 + n + 41$}{f(n) = n\^{}2 + n + 41}}
\label{L-Polynomfunktion01-41}
   This sequence starts off very {\em promisingly},  
   but is far from being a proof.
$$
 \begin{array}{lll}
f(0) = 41 & & \mapsto {\rm ~prime} \\
f(1) = 43 & & \mapsto {\rm ~prime} \\
f(2) = 47 & & \mapsto {\rm ~prime} \\
f(3) = 53 & & \mapsto {\rm ~prime} \\
f(4) = 61 & & \mapsto {\rm ~prime} \\
f(5) = 71 & & \mapsto {\rm ~prime} \\
f(6) = 83 & & \mapsto {\rm ~prime} \\
f(7) = 97 & & \mapsto {\rm ~prime} \\
\vdots \\
f(33) = 1,163 & & \mapsto {\rm ~prime} \\
f(34) = 1,231 & & \mapsto {\rm ~prime} \\
f(35) = 1,301 & & \mapsto {\rm ~prime} \\
f(36) = 1,373 & & \mapsto {\rm ~prime} \\
f(37) = 1,447 & & \mapsto {\rm ~prime} \\
f(38) = 1,523 & & \mapsto {\rm ~prime} \\
f(39) = 1,601 & & \mapsto {\rm ~prime} \\
f(40) = 1681 & = 41 \cdot 41 & \mapsto {\rm ~NOT~prime}! \\
f(41) = 1763 & = 41 \cdot 43 & \mapsto {\rm ~NOT~prime}! \\
\end{array} 
$$

@@

% --------------------------------------------------------------------------
\subsection{\texorpdfstring{$f(n) = n^2 - 79 \cdot n + 1,601$}{f(n) = n\^{}2 - 79 ... n + 1,601}}
\label{L-Polynomfunktion02-1601}   \label{SrcPrimArith1}
   This function\footnote{%
   See \hyperlink{Primes_Appendix_A}{Appendix A of this chapter} for the source code
   to compute the table using Sage\index{Sage}.}  %Pari-GP.\index{Pari-GP}}
   delivers prime numbers for all values from $n=0$ to $n=79$.
Unfortunately $f(80) = 1,681 = 11 \cdot 151$ is not a prime number. To this
date, no function has been found that delivers more prime numbers in a row. On
the other hand, each prime occurs twice (first in the decreasing then in the
increasing sequence), which means that the algorithm delivers a total of 40
different prime values (these are the same ones as delivered by the function
in chapter~\ref{L-Polynomfunktion01-41})\footnote{%
   Another quadratic polynom, which delivers these primes, is: 
   $f(n) = n^2 - 9 \cdot n + 61$.\\
   Among the first 1000 sequence elements more than 50\% are prime
   (See \hyperlink{Primes_Appendix_A}{Appendix A of this chapter}).}.

$$
\begin{array}{|ll||ll|}
\hline
f(0)  = 1.601 & \mapsto {\rm ~prim}   &   f(26) =   223 & \mapsto {\rm ~prim} \\
f(1)  = 1.523 & \mapsto {\rm ~prim}   &   f(27) =   197 & \mapsto {\rm ~prim} \\
f(2)  = 1.447 & \mapsto {\rm ~prim}   &   f(28) =   173 & \mapsto {\rm ~prim} \\
f(3)  = 1.373 & \mapsto {\rm ~prim}   &   f(29) =   151 & \mapsto {\rm ~prim} \\
f(4)  = 1.301 & \mapsto {\rm ~prim}   &   f(30) =   131 & \mapsto {\rm ~prim} \\
f(5)  = 1.231 & \mapsto {\rm ~prim}   &   f(31) =   113 & \mapsto {\rm ~prim} \\
f(6)  = 1.163 & \mapsto {\rm ~prim}   &   f(32) =    97 & \mapsto {\rm ~prim} \\
f(7)  = 1.097 & \mapsto {\rm ~prim}   &   f(33) =    83 & \mapsto {\rm ~prim} \\
f(8)  = 1.033 & \mapsto {\rm ~prim}   &   f(34) =    71 & \mapsto {\rm ~prim} \\
f(9)  =   971 & \mapsto {\rm ~prim}   &   f(35) =    61 & \mapsto {\rm ~prim} \\
f(10) =   911 & \mapsto {\rm ~prim}   &   f(36) =    53 & \mapsto {\rm ~prim} \\
f(11) =   853 & \mapsto {\rm ~prim}   &   f(37) =    47 & \mapsto {\rm ~prim} \\
f(12) =   797 & \mapsto {\rm ~prim}   &   f(38) =    43 & \mapsto {\rm ~prim} \\
f(13) =   743 & \mapsto {\rm ~prim}   &   f(39) =    41 & \mapsto {\rm ~prim} \\
f(14) =   691 & \mapsto {\rm ~prim}   &   f(40) =    41 & \mapsto {\rm ~prim} \\              
f(15) =   641 & \mapsto {\rm ~prim}   &   f(41) =    43 & \mapsto {\rm ~prim} \\              
f(16) =   593 & \mapsto {\rm ~prim}   &   f(42) =    47 & \mapsto {\rm ~prim} \\              
f(17) =   547 & \mapsto {\rm ~prim}   &   f(43) =    53 & \mapsto {\rm ~prim} \\              
f(18) =   503 & \mapsto {\rm ~prim}   &   \cdots  &  \\                                       
f(19) =   461 & \mapsto {\rm ~prim}   &   f(77) = 1.447 & \mapsto {\rm ~prim} \\             
f(20) =   421 & \mapsto {\rm ~prim}   &   f(78) = 1.523 & \mapsto {\rm ~prim} \\             
f(21) =   383 & \mapsto {\rm ~prim}   &   f(79) = 1.601 & \mapsto {\rm ~prim} \\             
f(22) =   347 & \mapsto {\rm ~prim}   &   f(80) = 41 \cdot 41 & \mapsto {\rm ~NOT~prim!} \\
f(21) =   383 & \mapsto {\rm ~prim}   &   f(81) = 41 \cdot 43 & \mapsto {\rm ~NOT~prim!} \\ 
f(22) =   347 & \mapsto {\rm ~prim}   &   f(82) = 1.847 & \mapsto {\rm ~prim} \\             
f(23) =   313 & \mapsto {\rm ~prim}   &   f(83) = 1.933 & \mapsto {\rm ~prim} \\              
f(24) =   281 & \mapsto {\rm ~prim}   &   f(84) = 43 \cdot 47 &  \mapsto {\rm ~NOT~prim!} \\
f(25) =   251 & \mapsto {\rm ~prim}   &   & \\
\hline
\end{array} 
$$

% --------------------------------------------------------------------------
\vskip +20 pt
\subsection[Polynomial functions 
    \texorpdfstring{$f(x) = a_n x^n + a_{n-1}x^{n-1} + \cdots + a_1 x^1 + a_0$}
                   {f(x) = a\_n x\^{}n + a\_n-1 x\^{}n-1 + ... + a\_1 x\^{}1 + a\_0}]
    {Polynomial functions\index{Polynomial} 
    $f(x) = a_n x^n + a_{n-1}x^{n-1} + \cdots + a_1 x^1 + a_0$ ~ 
    ($a_i$ in ${\mathbb Z}$, $n \geq 1$)}\index{Polynomial}

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\subsection[Catalan's conjecture]{Catalan's conjecture\footnotemark}
    \footnotetext{%
Eugene Charles Catalan, Belgian mathematician, May 5, 1814$-$Feb 14, 1894.\\
After him the so-called {\em Catalan numbers}
$A(n) = (1 / (n+1) ) * (2n)! / (n!)^2 $ \\
$=  1, 2, 5, 14, 42, 132, 429, 1430, 
4862, 16796, 58786, 208012, 742900, 2674440, 9694845, ... $
are named.
    }
\nopagebreak
Catalan \index{Catalan Eugene}  \index{Number!Catalan}
conjectured that $ C_4 \;$ is a prime:
$$
 \begin{array}{l}
C_0 = 2, \\
C_1 = 2^{C_0} - 1,  \\
C_2 = 2^{C_1} - 1,  \\
C_3 = 2^{C_2} - 1, \\
C_4 = 2^{C_3} - 1, \cdots \\
\end{array}
$$
\begin{sloppypar}
(see \url{http://www.utm.edu/research/primes/mersenne.shtml} under Conjectures and
Unsolved Problems).
\end{sloppypar}

@@

\vskip +12pt
\noindent \textbf{Web links:}\\
\url{http://www.math.Princeton.EDU/~arbooker/nthprime.html} \\
\url{http://www.utm.edu/research/primes/notes/by_year.html}.

@@

% ---------------------------------------------------------------------------
\subsection{Check primality of integers generated by quadratic functions}
\index{Sage}

@@

\noindent The last two lines of the output contain a small statistics.
You can see that $f(n)$ generates
48 primes when $0 \leq n \leq 50$, which is approximately 94\% of the
values generated by $f(n)$.\\
